Exact expressions for double descent and implicit regularization \\ via surrogate random design

### 2022-11-23 design/aistats/bayesian_experiments-appendix.tex
Working compile

This section presents additional details and experimental results omitted from
the main body of the paper. In addition to the \texttt{mg\_scale} dataset presented in
Section~\ref{s:experiments}, we also benchmarked on three other data sets
described in Table~\ref{tab:libsvm-datasets}.
\begin{table}[ht]
    \centering
    \caption{Datasets used in the experiments \citep{libsvm}.}
    \label{tab:libsvm-datasets}
    \begin{tabular}[t]{lcccc}
        \toprule
        & \texttt{mg\_scale} & \texttt{bodyfat\_scale} & \texttt{mpg\_scale} & \texttt{housing\_scale} \\
        \midrule
        $n$ & 1385 & 252 & 392 & 506 \\
        $d$ & 6 & 14 & 7 & 13\\
        \bottomrule
    \end{tabular}
\end{table}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/obj_grid.pdf}
    \caption{A-optimality values achieved by the methods compared. In all cases
        considered, we found our method (without SDP) to be superior to independent
        sampling methods like uniform and predictive length sampling. After paying the price
        to solve an SDP, our method (with SDP) is able to consistently match the performance
        of a greedy method which has been noted
        \citep{chamon2017approximate} to work well empirically.}
    \label{f:obj-grid}
\end{figure}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/runtime_grid.pdf}
    \caption{Runtimes of the methods compared. Our method (without SDP) is
        within an order of magnitude of greedy bottom up and faster in 3 out of
        4 cases. The gap between our method with and without SDP is
        attributable to the SDP solver, making investigation of more efficient
        solvers and approximate solutions an interesting direction for future
        work.
    }
    \label{f:runtimes}
\end{figure}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/ratios_grid.pdf}
    \caption{The ratio controlled by Lemma~\ref{l:guarantees}. This ratio converges
        to $1$ as $k \to n$ and is close to $1$ across all
        real world datasets,
        suggesting that $f_{\A}(\frac{k}{n} \Sigmab_\X)$
        is an appropriate problem-dependent scale for Bayesian A-optimal
        experimental design.
    }
    \label{f:ratios}
\end{figure}

>>>>

% This section presents additional details and experimental results omitted from
% the main body of the paper. 
In addition to the \texttt{mg\_scale} dataset presented in
Section~\ref{s:experiments}, we also benchmarked on three other data sets
described in Table~\ref{tab:libsvm-datasets}.
\begin{table}[ht]
  \centering
  \caption{Datasets used in the experiments \citep{libsvm}.}
  \label{tab:libsvm-datasets}
  \begin{tabular}[t]{lcccc}
    \toprule
        & \texttt{mg\_scale} & \texttt{bodyfat\_scale} & \texttt{mpg\_scale} & \texttt{housing\_scale} \\
    \midrule
    $n$ & 1385               & 252                     & 392                 & 506                     \\
    $d$ & 6                  & 14                      & 7                   & 13                      \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{figure}[htpb]
  \centering
  \includegraphics[width=\textwidth]{design/bayesian_figures/obj_grid.pdf}
  \caption{A-optimality values achieved by the methods compared. In all cases
    considered, we found our method (without SDP) to be superior to independent
    sampling methods like uniform and predictive length sampling. After paying the price
    to solve an SDP, our method (with SDP) is able to consistently match the performance
    of a greedy method which has been noted
    \citep{chamon2017approximate} to work well empirically.}
  \label{f:obj-grid}
\end{figure}

@@

\begin{figure}[htpb]
  \centering
  \includegraphics[width=\textwidth]{design/bayesian_figures/runtime_grid.pdf}
  \caption{Runtimes of the methods compared. Our method (without SDP) is
    within an order of magnitude of greedy bottom up and faster in 3 out of
    4 cases. The gap between our method with and without SDP is
    attributable to the SDP solver, making investigation of more efficient
    solvers and approximate solutions an interesting direction for future
    work.
  }
  \label{f:runtimes}
\end{figure}

@@

\begin{figure}[htpb]
  \centering
  \includegraphics[width=\textwidth]{design/bayesian_figures/ratios_grid.pdf}
  \caption{The ratio controlled by Lemma~\ref{l:guarantees}. This ratio converges
    to $1$ as $k \to n$ and is close to $1$ across all
    real world datasets,
    suggesting that $f_{\A}(\frac{k}{n} \Sigmab_\X)$
    is an appropriate problem-dependent scale for Bayesian A-optimal
    experimental design.
  }
  \label{f:ratios}
\end{figure}

### 2022-11-23 design/continuous_nips/continuous-empirical.tex
Updates from Overleaf
Working compile

\begin{figure}[t]
  \includegraphics[width=\textwidth]{continuous_figures/decays.pdf}
  \vspace{-1cm}
  \caption{Scree-plots of $\Sigmab$ for the eigenvalue decays examined
    in our empirical valuations.
    % Here $d=100$ for visualization, whereas
    % our experiments increase $d$ while preserving the ratio $n/d$ and
    % the decay profile,
    % with $\lambda_{\text{max}}(\Sigmab) = 1$ to
    % $\lambda_{\text{min}}(\Sigmab) = 10^{-4}$.
  }
  \label{fig:eig-decays}
\end{figure}

@@

% Theorem \ref{t:asymptotic} shows that the surrogate MSE expressions
% derived in Theorem \ref{t:mse} are asymptotically consistent with the
% true MSE for the i.i.d.~design under a class of sub-gaussian data
% distributions.
In this section, we empirically quantify the convergence rates for
the asymptotic result of Theorem~\ref{t:asymptotic}.
% when $\mu$ is a centered multivariate% Gaussian $\Nc(\zero,\Sigmab).$
We focus on the under-determined regime (i.e.,
$n<d$) and separate the evaluation into the bias and
variance terms, following the MSE decomposition given
in \eqref{eq:mse-derivation}. Consider  $\X = \Z\Sigmab^{1/2} $, where the entries of $\Z$ are
i.i.d. standard Gaussian, and define:\vspace{-1mm}
\begin{enumerate}
  \item Variance discrepancy:\quad
    $\big|\frac{\E[\tr((\X^\top\X)^\dagger)]}{\Vc(\Sigmab,n)}-1\big|$ where
    $\Vc(\Sigmab,n)=\frac{1-\alpha_n}{\lambda_n}$.
  \item Bias discrepancy:\quad
     $\sup_{\w\in\R^d\backslash\{\zero\}}\big|\frac{\w^\top\E[\I-\X^\dagger\X]\w}
     {\w^\top\Bc(\Sigmab,n)\w} - 1\big|$
    where $\Bc(\Sigmab,n) = \lambda_n(\Sigmab+\lambda_n\I)^{-1}$.
  \end{enumerate}\vspace{-1mm}
   Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
so our surrogate MSE can be written as
$\Mc=\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$, and when both
discrepancies are bounded by $\epsilon$, then $(1-2\epsilon)\Mc\leq\MSE{\X^\dagger\y}\leq (1+2\epsilon)\Mc$.
In our experiments, we consider four standard eigenvalue decay profiles
for $\Sigmab$, including polynomial and exponential decay (see
 \Cref{fig:eig-decays} and \Cref{sec:eig-decay-details}).
\begin{figure} %[ht]
  \includegraphics[width=\textwidth]{continuous_figures/wishart_var.pdf}
  \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
  \vspace{-.8cm}
  \caption{
    Empirical verification of the asymptotic consistency of surrogate MSE.
    We show the discrepancies for the variance (top) and bias
    (bottom),  with bootstrapped $95\%$ confidence intervals, as $d$
    increases and $n/d$ is fixed. We observe
     $O(1/d)$ decay (linear with slope $-1$ on a log-log plot).
  }
  \label{f:conj-wishart}
\end{figure}

Figure~\ref{f:conj-wishart} (top) plots the variance discrepancy (with
$\E[\tr((\X^\top\X)^\dagger)]$ estimated via Monte Carlo
sampling and bootstrapped confidence intervals) as $d$ increases from $10$ to
$1000$, across a range of aspect ratios $n/d$. In all cases, we observe that
the discrepancy decays to zero at a rate of $O(1/d)$.
Figure~\ref{f:conj-wishart} (bottom) plots the bias discrepancy, with the same
rate of decay observed throughout.  Note that the range of $d$ is smaller than
in Figure \ref{f:conj-wishart} (top) because the large number of Monte Carlo
samples (up to two million) required for this experiment made the computations
much more expensive (more details in Appendix \ref{a:empirical}). Based on the
above empirical results, we conclude with a conjecture.
\begin{conjecture}
  \label{c:1-over-d-rate}
  When $\mu$ is a centered multivariate Gaussian and its covariance
  has a constant condition
  number, then, for $n/d$ fixed, the surrogate MSE satisfies:
  $\big|\frac{\textnormal{MSE}[\X^\dagger\y]}{\Mc}-1\big|= O(1/d)$.
\end{conjecture}

@@

>>>>

\begin{figure}[t]
  \includegraphics[width=\textwidth]{design/continuous_figures/decays.pdf}
  \vspace{-1cm}
  \caption{Scree-plots of $\Sigmab$ for the eigenvalue decays examined
    in our empirical valuations.
    % Here $d=100$ for visualization, whereas
    % our experiments increase $d$ while preserving the ratio $n/d$ and
    % the decay profile,
    % with $\lambda_{\text{max}}(\Sigmab) = 1$ to
    % $\lambda_{\text{min}}(\Sigmab) = 10^{-4}$.
  }
  \label{fig:eig-decays}
\end{figure}

@@

% Theorem \ref{t:asymptotic} shows that the surrogate MSE expressions
% derived in Theorem \ref{t:mse} are asymptotically consistent with the
% true MSE for the i.i.d.~design under a class of sub-gaussian data
% distributions.
In this section, we empirically quantify the convergence rates for
the asymptotic result of Theorem~\ref{t:asymptotic}.
% when $\mu$ is a centered multivariate% Gaussian $\Nc(\zero,\Sigmab).$
We focus on the under-determined regime (i.e.,
$n<d$) and separate the evaluation into the bias and
variance terms, following the MSE decomposition given
in \eqref{eq:mse-derivation}. Consider  $\X = \Z\Sigmab^{1/2} $, where the entries of $\Z$ are
i.i.d. standard Gaussian, and define:\vspace{-1mm}
\begin{enumerate}
  \item Variance discrepancy:\quad
        $\big|\frac{\E[\tr((\X^\top\X)^\dagger)]}{\Vc(\Sigmab,n)}-1\big|$ where
        $\Vc(\Sigmab,n)=\frac{1-\alpha_n}{\lambda_n}$.
  \item Bias discrepancy:\quad
        $\sup_{\w\in\R^d\backslash\{\zero\}}\big|\frac{\w^\top\E[\I-\X^\dagger\X]\w}
          {\w^\top\Bc(\Sigmab,n)\w} - 1\big|$
        where $\Bc(\Sigmab,n) = \lambda_n(\Sigmab+\lambda_n\I)^{-1}$.
\end{enumerate}\vspace{-1mm}
Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
so our surrogate MSE can be written as
$\Mc=\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$, and when both
discrepancies are bounded by $\epsilon$, then $(1-2\epsilon)\Mc\leq\MSE{\X^\dagger\y}\leq (1+2\epsilon)\Mc$.
In our experiments, we consider four standard eigenvalue decay profiles
for $\Sigmab$, including polynomial and exponential decay (see
\Cref{fig:eig-decays} and \Cref{sec:eig-decay-details}).
\begin{figure} %[ht]
  \includegraphics[width=\textwidth]{design/continuous_figures/wishart_var.pdf}
  \includegraphics[width=\textwidth]{design/continuous_figures/proj_bias.pdf}
  \vspace{-.8cm}
  \caption{
    Empirical verification of the asymptotic consistency of surrogate MSE.
    We show the discrepancies for the variance (top) and bias
    (bottom),  with bootstrapped $95\%$ confidence intervals, as $d$
    increases and $n/d$ is fixed. We observe
    $O(1/d)$ decay (linear with slope $-1$ on a log-log plot).
  }
  \label{f:conj-wishart}
\end{figure}

Figure~\ref{f:conj-wishart} (top) plots the variance discrepancy (with
$\E[\tr((\X^\top\X)^\dagger)]$ estimated via Monte Carlo
sampling and bootstrapped confidence intervals) as $d$ increases from $10$ to
$1000$, across a range of aspect ratios $n/d$. In all cases, we observe that
the discrepancy decays to zero at a rate of $O(1/d)$.
Figure~\ref{f:conj-wishart} (bottom) plots the bias discrepancy, with the same
rate of decay observed throughout.  Note that the range of $d$ is smaller than
in Figure \ref{f:conj-wishart} (top) because the large number of Monte Carlo
samples (up to two million) required for this experiment made the computations
much more expensive (more details in \Cref{a:empirical}). Based on the
above empirical results, we conclude with a conjecture.
\begin{conjecture}
  \label{c:1-over-d-rate}
  When $\mu$ is a centered multivariate Gaussian and its covariance
  has a constant condition
  number, then, for $n/d$ fixed, the surrogate MSE satisfies:
  $\big|\frac{\textnormal{MSE}[\X^\dagger\y]}{\Mc}-1\big|= O(1/d)$.
\end{conjecture}

@@

\subsection{Additional details for empirical evaluation}
\label{a:empirical}

Our empirical investigation of the rate of asymptotic convergence in \Cref{t:asymptotic} (and, more
specifically, the variance and bias discrepancies defined in
Section~\ref{sec:asymp-conj-details}), in the context of
Gaussian random matrices, is related to open problems which have been
extensively studied in the literature. Note that when
$\X=\Z\Sigmab^{1/2}$ were $\Z$ has i.i.d.~Gaussian entries (as in
Section \ref{sec:asymp-conj-details}), then $\W=\X^\top\X$ is known as the
pseudo-Wishart distribution (also called the singular Wishart),
denoted as $\W \sim \Pc\Wc(\Sigmab, n)$, and the variance
term from the MSE can be written as $\sigma^2\E[\tr(\W^\dagger)]$.
\cite{srivastava2003} first derived the probability density function of the
pseudo-Wishart distribution, and
\cite{cook2011} computed the first and second moments of generalized
inverses. However, for the Moore-Penrose inverse and arbitrary
covariance $\Sigmab$, \cite{cook2011} claims that the quantities required to
express the mean ``do not have tractable closed-form representation.''
The bias term, $\w^{*\top}\E[\I-\X^\dagger\X]\w^*$, has connections to
directional statistics.  Using the SVD,
we have the equivalent representation $\X^\dagger \X = \V \V^\top$ where $\V$
is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal $n$-frames
in $\R^d$).  The distribution of $\V$ is known as the matrix angular central
Gaussian (MACG) distribution \citep{chikuse1990matrix}. While prior work has
considered high dimensional limit theorems \citep{CHIKUSE1991145} as well as
density estimation and hypothesis testing \citep{CHIKUSE1998188} on $V_{n,d}$,
they only analyzed the invariant measure (which corresponds in our setting to
$\Sigmab = \I$), and to our knowledge a closed form expression of
$\E[\V\V^\top]$ where $\V$ is distributed according to MACG with arbitrary
$\Sigmab$ remains an open question.

For analyzing the rate of decay of variance and bias discrepancies (as
defined in Section \ref{sec:asymp-conj-details}), it suffices to only consider diagonal
covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D \Q^\top$ is
its eigendecomposition and $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Q\D\Q^\top)$,
then we have for $\W \sim \Pc\Wc(\Sigmab, n)$ that $\W \overset{d}{=} \X^\top
  \X$ and hence, defining $\Xt\sim\Nc_{n,d}(\zero,\I_n \otimes \D)$, by linearity
and unitary invariance of trace,
\begin{align*}
  \E[\tr(\W^\dagger)]
   & = \tr\big( \E[(\X^\top\X)^\dagger] \big)
  = \tr\Big( \Q\E\big[(\Xt^\top\Xt)^\dagger\big]\Q^\top \Big)
  = \tr\Big( \E\big[(\Xt^\top\Xt)^\dagger\big] \Big)
  = \E\left[\tr \big((\Xt^\top\Xt)^\dagger\big) \right].
\end{align*}
Similarly, we have that $\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$,
and a simple calculation shows that the bias discrepancy is
also independent of the choice of matrix $\Q$.

In our experiments, we increase $d$ while keeping the aspect ratio $n/d$
fixed and examining the rate of decay of the discrepancies.
We estimate $\E\big[\tr(\W^\dagger)\big]$ (for the variance) and
$\E[\I-\X^\dagger\X]$ (for the bias) through Monte Carlo sampling.
Confidence intervals are constructed using ordinary bootstrapping for
the variance. We rewrite the supremum over $\w$ in bias discrepancy as
a spectral norm:
\[\big\|\Bc(\Sigmab,n)^{-\frac12}\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-\frac12} -
  \I\big\|,\]
and apply existing methods for constructing bootstrapped operator
norm confidence intervals described in \cite{lopes2019bootstrapping}.  To
ensure that estimation noise is sufficiently small, we continually increase the
number of Monte Carlo samples until the bootstrap confidence intervals are
within $\pm 12.5\%$ of the measured discrepancies.  We found that while
variance discrepancy required a relatively small number of trials (up
to one thousand), estimation noise was much larger for the bias
discrepancy, and it necessitated over two million trials to obtain
good estimates near $d=100$.

\subsection{Eigenvalue decay profiles}
\label{sec:eig-decay-details}

Letting $\lambda_i(\Sigmab)$ be the $i$th largest eigenvalue of
$\Sigmab$, we consider the
following eigenvalue profiles (visualized in Figure~\ref{fig:eig-decays}):
\begin{itemize}
  \item \texttt{diag\_linear}: linear decay, $\lambda_i(\Sigmab)= b-a i$;
  \item \texttt{diag\_exp}: exponential decay, $\lambda_i(\Sigmab) = b\,10^{- a i} $;
  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_i(\Sigmab) = (b-a i)^2$;
  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i(\Sigmab) = b i^{-a}$.
\end{itemize}
The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
$\kappa(\Sigmab) = 10^{4}$ remains constant).
